Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2015-05-14 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Adding Bench Exp results

\Supervisor{Dr Charles Fleischmann}
\Department{Department of Fire Safety Engineering}

@@

For piloted ignition tests, an electric spark ignitor is positioned at the top of vertical samples and over the center of horizontal samples.  Combustion products and dilution air are extracted through the hood and exhaust duct by a high temperature fan. The flow rate can be adjusted between 0.01 and 0.03~m$^3$/s. The volumetric flow rate is kept constant during testing. The sample is mounted on a load cell to determine mass loss rate during a test, and smoke obscuration is measured using a laser light source. The gas flow rate in the exhaust duct is calculated from the pressure drop across and temperature at an orifice plate in the duct. Finally, the concentrations of oxygen, carbon dioxide, carbon monoxide, were measured using appropriate instruments. Heat release rate is calculated from the gas concentration and mass flow measurements.

For this study samples were cut from painted gypsum wallboard panels. The samples were 100~mm square, and 12.7~mm thick. The experiments were conducted with the samples in the horizontal position.  The key measurements and observations made with the cone calorimeter were heat release rate per unit area of the samples and time to ignition. These experiments were conducted with heat fluxes of 25, 35, 50, 70 and 75 kW/m$^2$. Several other cone calorimeter experiments were conducted with a single type K thermocouple, installed in a slit in the painted paper on the gypsum wallboard samples.  This was the same manner of thermocouple installation used in the instrumented gypsum board wall experiments.  These experiments were conducted at heat fluxes of 25, 30, and 35 kW/m$^2$ to determine ignition time and surface temperature pairs.

@@

A pilot flame from a 254~mm long wide stainless steel burner tube, as specified in ASTM~648, was used to ignite the test sample near the edge of the sample closet to the panel~\cite{ASTM_E648}. The objective of these experiments was to determine the critical radiant flux for the painted gypsum wallboard samples.  The critical radiant flux corresponds to the flux at the point of maximum flame propagation, or the point on the sample where the flame spread stopped. For this study the samples examined were cut from painted gypsum wallboard panels that were prepared from the supply of panels that were used in the compartment experiments.

@@

For the purposes of this study a modified version of the test apparatus was used.  The radiant panel and the radiant panel controls remained in place as documented in ASTM~E162. However the sample mounting system was changed. A rail and trolley system was developed by Lawson and Twilley to expose fire fighter protective equipment to various heat fluxes by being able to move the test specimen to a predetermined position that correlated with a given heat flux.  This trolley was used to hold a sample of painted gypsum wallboard that was 362~mm square.  The samples were vertical and parallel to the radiant panel. Figure~\ref{Rad_Panel} has photographs which show the radiant panel test arrangement.

\begin{figure}
	\centering
	\includegraphics[width=2.92in]{../Figures/RAD_pan}
	\includegraphics[width=2.08in]{../Figures/RAD_pan_iso} \\
	\caption[Photographs of the radiant panel experimental arrangement]{Photographs of the radiant panel experimental arrangement, on the left is a photograph which shows the radiant panel on the right, the aluminum heat shield in the center and the moveable mounting frame for the gypsum wallboard sample on the left.  The photograph on the right shows an isometric view of the test arrangement.}
	\label{Rad_Panel}
\end{figure}

Heat flux gauge measurements, taken at the center position of the sample, were used to determine positions on the trolley track for a specific radiant exposure. Positions were marked for 10, 12.5, 15, 17.5, 20, 22.5, and 25~kW/m$^2$. The area surrounding the center of the maintained the desired heat flux within at least 70~mm on either side of the center point and 100~mm above and below the center point.  In other words, the area around the center point of the sample had heat flux measures similar, within the uncertainty of the device, to the center point measurement. Outside of the central area of the samples, the heat flux would decrease as the position approached the edge of the sample.  The center point heat flux measures vs the distance from the radiating surface is provided in Figure~\ref{Rad_Panel_Cal}

To conduct the experiments, the radiant panel was ignitied and allowed to pre-heat for at least 60 minutes.  The water cooled, Schmidt-Boelter total heat flux gauge was mounted in the center the sample holder, and the heat flux positions were confirmed.  Just before the sample was mounted on the movable trolley, an aluminium heat shield was placed between the radiant panel and the sample holder.  Once the sample was mounted and the sample was moved to the appropriate distance from the radiant panel based on the heat flux desired for the experiment, the heat shield was removed that the experiment began.  Each sample was exposed for 600 seconds. Each sample had a single bare-bead, Chromel-Alumel (type K) thermocouple, with a 0.5~mm nominal diameter bead in contact with the center point of each sample. During the experiments the samples were monitored for temperature and ignition.    examined and area of thermal damage to the painted paper surface of the gypsum board was compared with heat flux mapping.  This was another means of potentially obtaining a measure of the critical heat flux. .11

@@

\subsection{Ignition Temperature}

@@

>>>>

\Supervisor{Dr Charles Fleischmann}
\Department{College of Engineering, Civil and Natural Resources Engineering}

@@

For the ignition tests, an electric spark ignitor was positioned over the center of the horizontal samples to serve as a pilot.  Combustion products and dilution air were extracted through the hood and exhaust duct by a high temperature fan. The flow rate can be adjusted between 0.01 and 0.03~m$^3$/s. The volumetric flow rate was kept constant during testing. The sample was mounted on a load cell to determine mass loss rate during a test, and smoke obscuration was measured using a laser light source. The gas flow rate in the exhaust duct was calculated from the pressure drop across and temperature at an orifice plate in the duct. Finally, the concentrations of oxygen, carbon dioxide, carbon monoxide, were measured using appropriate instruments. Heat release rate was calculated from the gas concentration and mass flow measurements.

For this study samples were cut from painted gypsum wallboard panels. The samples were 100~mm square and 12.7~mm thick. The experiments were conducted with the samples in the horizontal position.  The key measurements and observations made with the cone calorimeter were heat release rate per unit area of the samples and time to ignition. These experiments were conducted with heat fluxes of 25, 35, 50, 70 and 75 kW/m$^2$. Several other cone calorimeter experiments were conducted with a single type K thermocouple, installed in a slit in the painted paper on the gypsum wallboard samples.  This was the same manner of thermocouple installation used in the instrumented gypsum board wall experiments.  These experiments were conducted at heat fluxes of 30, 35, and 50 kW/m$^2$ to determine ignition time and surface temperature pairs.

@@

A diffusion, propane-fueled, pilot flame from a 254~mm long wide stainless steel burner tube, as specified in ASTM~648, was used to ignite the test sample near the edge of the sample closet to the panel~\cite{ASTM_E648}. The objective of these experiments was to determine the critical radiant flux for the painted gypsum wallboard samples.  The critical radiant flux corresponds to the flux at the point of maximum flame propagation, or the point on the sample where the flame spread stopped. For this study the samples examined were cut from painted gypsum wallboard panels that were prepared from the supply of panels that were used in the compartment experiments.

@@

For the purposes of this study a modified version of the test apparatus was used.  The radiant panel and the controls for the radiant panel  remained in place as documented in ASTM~E162. However the sample mounting system was changed. A rail and trolley system was developed by Lawson and Twilley to expose fire fighter protective equipment to various heat fluxes by being able to move the test specimen to a predetermined position that correlated with a given heat flux~\cite{Lawson:1999}.  This trolley was used to hold a sample of painted gypsum wallboard that was 362~mm square.  The samples were vertical and parallel to the radiant panel. Figure~\ref{Rad_Panel} has photographs which show the radiant panel test arrangement.

\begin{figure}
	\centering
	\includegraphics[width=2.92in]{../Figures/RAD_pan}
	\includegraphics[width=2.08in]{../Figures/RAD_pan_iso} \\
	\caption[Photographs of the radiant panel experimental arrangement]{Photographs of the radiant panel experimental arrangement, on the left is a photograph which shows the radiant panel on the right, the aluminum heat shield in the center and the moveable sample mounting frame trolley for the gypsum wallboard sample on the left.  The photograph on the right shows an isometric view of the test arrangement.}
	\label{Rad_Panel}
\end{figure}

Prior to conducting experiments with the gypsum wallboard samples, heat flux gauge measurements taken at the center position of the sample were used to determine the positions on the trolley track for a specific radiant exposure. The positions were marked for 10, 12.5, 15, 17.5, 20, 22.5, and 25~kW/m$^2$. The area surrounding the center of the maintained the desired heat flux within at least 70~mm on either side of the center point and 100~mm above and below the center point.  In other words, the area around the center point of the sample had heat flux measures similar, within the uncertainty of the device, to the center point measurement. Outside of the central area of the samples, the heat flux would decrease as the position approached the edge of the sample.  The center point heat flux measures vs the distance from the radiating surface is provided in Figure~\ref{Rad_Panel_Cal}

To conduct the experiments, the radiant panel was ignited and allowed to pre-heat for at least 60 minutes.  The water cooled, Schmidt-Boelter total heat flux gauge was mounted in the center the sample holder, and the heat flux positions were confirmed. The heat flux gauge was then removed from the sample holder trolley. Just before the sample was mounted on the movable trolley, an aluminum heat shield was placed between the radiant panel and the sample holder trolley.  Once the sample was mounted and the sample was moved to the appropriate distance from the radiant panel based on the heat flux desired for the experiment, the heat shield was removed.  The removal of the heat shield would be recorded as the beginning of the exposure and therefore the beginning of the experiment.   Each sample was exposed for 600 seconds. Each sample had a single bare-bead, Chromel-Alumel (type K) thermocouple, with a 0.5~mm nominal diameter bead in contact with the center point of each sample. During the experiments the samples were monitored for temperature and ignition. After the exposure time was completed the heat shield was once again placed between the sample and the radiant panel.  The samples were cooled, removed and examined.  The area of thermal damage to the painted paper surface of the gypsum board was compared with exposure heat flux.  This was another means of potentially obtaining a measure of the critical heat flux.

@@

\subsection{Ignition and Ignition Temperature}

Three replicate time to ignition experiments were conducted with the cone calorimeter apparatus at each heat flux.  The results of these experiments are shown in Table~\ref{tab:Gypsum_wallboard_timetoign}.  The time to ignition values in the table are the average of the three times.  Note that no ignition occurred at 25~kg/m$^2$.  The exposure at 25~kg/m$^2$ resulted in charring and cracking of the painted paper surface of the gypsum wallboard sample.  This began to occur, on average, at 270 s of exposure time. The samples were exposed to 25~kg/m$^2$ for 1800~s each and no ignition occurred.  In the experiments with heat fluxes of 35~kg/m$^2$ or higher, ignition did occur and the painted paper surface of the gypsum board was consumed, leaving the gypsum board exposed.  There were no surprises in the data with regard to the heat flux exposures.  The time to ignition was reduced as the incident heat flux was increased.

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline Incident Heat Flux & Ignition & Time to Ignition &  Fire Effect  \\
		\hline (kg/m$^2$) &  & (s)  &   \\
		\hline 25	& No 	& NA 		& Paint \& paper charred \& cracked \\
		\hline 35 	& Yes 	& 73	 	& Paint \& paper consumed \\
		\hline 50	& Yes 	& 35 	 	& Paint \& paper consumed   \\
		\hline 70	& Yes 	& 21 	 	& Paint \& paper consumed \\
		\hline 75	& Yes 	& 17 		& Paint \& paper consumed \\
		\hline
	\end{tabular}
	\caption[Painted gypsum wallboard time to ignition results]{Painted gypsum wallboard time to ignition results from the cone calorimeter}
	\label{tab:Gypsum_wallboard_timetoign}
\end{table}

Additional cone calorimeter experiments were conducted in the University of Canterbury Fire Laboratory.  In these experiments the purpose was to obtain an estimate of the temperature of the painted paper surface at the time of ignition.   Samples of painted gypsum wallboard were shipped from the NIST Fire Research Laboratory in Gaithersburg, MD., USA.  In these experiments an additional incident heat flux level of 30~kg/m$^2$ was added to provide another test between the 25 and 35~kg/m$^2$ levels.  The averaged ignition time and temperature pairs are shown in Table~\ref{tab:Gypsum_wallboard_igntemp}.  The average time to ignition values between the two different calorimeters are within $\pm~50~\%$ and $\pm~10~\%$ at incident heat flux values of 35 and 50~kg/m$^2$, respectively.  The average ignition temperatures from the nine experiments ranged from 335 to 375~$^{\circ}$C.

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline Incident Heat Flux & Ignition & Time to Ignition & Ignition Temperature \\
		\hline (kg/m$^2$) & (Yes or No) & (s)  & Temperature ($^{\circ}$C)   \\
		\hline 30 	& Yes 	& 112	& 375 	 \\
		\hline 35	& Yes 	& 50 	& 335 	 \\
		\hline 50	& Yes 	& 38 	& 371 	 \\
		\hline
	\end{tabular}
	\caption[Painted gypsum wallboard ignition temperature results]{Painted gypsum wallboard ignition temperature results from exposure in the cone calorimeter}
	\label{tab:Gypsum_wallboard_igntemp}
\end{table}

In the radiant flooring panel experiments the gypsum wallboard was exposed to an incident heat flux that ranged from approximately 21~kg/m$^2$ to kg/m$^2$ over the length of the sample. In addition to the heat flux from the radiant panel, there is a small flame front approximately 10 to 30~mm in height that serves as the pilot along the length of the same and provides additional energy to the fuel immediately ahead of the flame.  The impact is to a very limited area ahead of the flame, estimated to be less than 5~mm, since the flame is drawn back to the leading edge of the sample (burned portion) by the draft of the radiant panel.  This effect is shown in Figure~\ref{FRP_flamefront}. In Figure~\ref{FRP_flamefront} several thermocouples can be seen. The exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0~mm nominal diameter, were installed by pulling the positive and negative wires through holes drilled through the gypsum wallboard sample and positioning the bead of the thermocouple in a slit in the paper face made with a razor blade.  This was the same installation method that was used in the instrumented gypsum board wall experiments and to measure the ignition temperature of the cone calorimeter samples.  The thermocouples that were installed in the center of the sample at 100~mm intervals beginning at 100~mm from the edge of the sample closest to the radiant panel.  Three replicate experiments were conducted.  The average time that the flame front reached the thermocouples and the temperature of the thermocouple just prior to the painted paper surface of the gypsum wallboard sample igniting is presented in Table~\ref{tab:FRP_timetemp}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/FRP_flamefront} \\
	\caption[Photograph of the flame front burning on the gypsum wallboard sample in the flooring radiant panel] {Photograph of the flame front burning on the gypsum wallboard sample in the flooring radiant panel. Thermocouples imbedded at 100~mm intervals in the surface of the painted paper can be seen in front of and behind the flame front. The flame is leaning back in the direction the radiant panel.}
	\label{FRP_flamefront}
\end{figure}

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|c|}
		\hline Postion & Incident Heat Flux & Ignition & Time to Ignition & Ignition Temperature \\
		\hline (mm) & (kg/m$^2$) & (Yes or No) & (s)  & Temperature ($^{\circ}$C)   \\
		\hline 100 	& 21	& Yes 	& 133	& 315 	 \\
		\hline 200	& 18 	& Yes 	& 198 	& 296 	 \\
		\hline 300	& 14	& Yes 	& 302 	& 279 	 \\
		\hline 400	& 10	& Yes 	& 502 	& 242 	 \\
		\hline 500	&  7	& No 	& NA 	&     	 \\
		\hline
	\end{tabular}
	\caption[Flooring radiant panel experiment flame spread results]{Flooring radiant panel experiment flame spread results including time to ignition at 100~mm intervals and surface temperature just prior to ignition.  The values are averages from three replicate experiments.}
	\label{tab:FRP_timetemp}
\end{table}

@@

\begin{table}
	\centering
	\begin{tabular}{|c|c|}
		\hline Incident Heat Flux & Heat Release Rate per Unit Area    \\
		\hline (kg/m$^2$) & (kg/m$^2$)  \\
		\hline 35 	& 163 	\\
		\hline 50	& 179   \\
		\hline 70	& 255  	\\
		\hline 75	& 231 	\\
		\hline
	\end{tabular}
	\caption[Painted gypsum wallboard time heat release rate per unit area results]{Painted gypsum wallboard heat release rate per unit area results from the cone calorimeter}
	\label{tab:Gypsum wallboard_HRRA}
\end{table}
